---
title: "Untitled"
author: "Nolan Moody (UCID 30044615)"
date: "2024-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

::: {style="margin-bottom:50px;"}
## Load Packages, Get Data

```{r}
install.packages("tidyverse") # Needed to pipe functions together
library("tidyverse") 
install.packages("mosaic")
library(mosaic)
install.packages("ggplot2")
library(ggplot2)
setwd("C:/Users/nolan/OneDrive/Desktop/DATA501 Project Files") 
load("./w - nhanes_1988_2018.RData")
```
:::

::: {style="margin-bottom:50px;"}
### Clean Age Data

```{r}
NHANES_AGE <- demographics_clean %>%
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier Data
         RIDAGEYR) # Recoded and cleaned age of all participants

# Change column name for easier processing
names(NHANES_AGE)[4] = "AGE"

# Convert age to numeric
NHANES_AGE$AGE <- as.numeric(NHANES_AGE$AGE)
```
:::

::: {style="margin-bottom:50px;"}
## Clean Gender Data

```{r}
NHANES_GENDER <- demographics_clean %>%
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier Data
         RIAGENDR) 	 # Recoded and cleaned gender for all participants

# Rename column for easier processing.
names(NHANES_GENDER)[4] <- "GENDER"

# Change categories. Male is originally measured as 1, female as 2. Change male to 0 and female to 1 to help with regression design.
NHANES_GENDER$GENDER <- ifelse(NHANES_GENDER$GENDER == 1, 0, 1)

# Change gender to categorical factor.
NHANES_GENDER$GENDER <- factor(NHANES_GENDER$GENDER)
```
:::

::: {style="margin-bottom:50px;"}
## Clean BMI Data

```{r}
NHANES_BMI_2 <- response_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         BMXBMI, # Adult BMI
         BMDBMIC) # Youth category	
# Categories for youth BMI:
# 1 = Underweight
# 2 = Normal Weight
# 3 = Overweight
# 4 = Obese

# Create an alternate measure of BMI that represents categories:
NHANES_BMI_2$BMI_CAT <- 
  ifelse(!is.na(NHANES_BMI_2$BMDBMIC), NHANES_BMI_2$BMDBMIC,
     ifelse(is.na(NHANES_BMI_2$BMXBMI), NA, 
            ifelse(NHANES_BMI_2$BMXBMI < 18.5, 1,
                    ifelse(NHANES_BMI_2$BMXBMI < 25, 2,
                           ifelse(NHANES_BMI_2$BMXBMI < 30, 3, 4)))))
# Change column name for convenience:
names(NHANES_BMI_2)[4] = "BMI"

# Subset to keep only data of interest for the primary BMI df:
NHANES_BMI <- NHANES_BMI_2 %>% 
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier variables
         BMI) # BMI as a number
# Ensure BMI is numeric
NHANES_BMI$BMI <- as.numeric(NHANES_BMI$BMI)

NHANES_BMI_alt <- NHANES_BMI_2 %>% 
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier variables
         BMI_CAT) # BMI category

```
:::

::: {style="margin-bottom:50px;"}
## Clean eGFR Data {style="margin-bottom:50px;"}

```{r}
NHANES_eGFR  <- response_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         VNEGFR) # Measured eGFR

# Rename variable for convenience
names(NHANES_eGFR)[4] = "eGFR"

# Ensure numerical measure
NHANES_eGFR$eGFR <- as.numeric(NHANES_eGFR$eGFR)
```
:::

::: {style="margin-bottom:50px;"}
## **Other Demographic Variables of Interest** {style="margin-bottom:50px;"}

INDFMIN2 Total family income (reported as a range value in dollars) 132610

INDFMPIR Poverty income ratio (PIR) - a ratio of family income to poverty threshold 122704

```{r}
NHANES_INCOME <- demographics_clean %>%
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier Data
         INDFMIN2, # Total Family Income
         INDFMPIR) # Poverty ratio
NHANES_INCOME$INDFMIN2 <- as.factor(NHANES_INCOME$INDFMIN2)
names(NHANES_INCOME)[4] = "INCOME_CAT"
names(NHANES_INCOME)[5] = "POVERTY_RATIO"
# Important to note: These two variables are highly correlated at ~0.85
```
:::

::: {style="margin-bottom:50px;"}
## Other Response Variables

```{r}
### Correlative and Causative
NHANES_RESPONSE_CORR <- response_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
          BMXWT, #	Weight (kg)
          BPXDI1, #	Diastolic: Blood pressure (first reading) mm Hg
          BPXML1, #	MIL: maximum inflation levels (mm Hg)
          BPXPLS, #	60 sec. pulse (30 sec. pulse * 2)
          LBDBANO, #	Basophils number (1000 cells/uL)
          LBDEONO, #	Eosinophils number (1000 cells/uL)
          LBDFOT, #	Serum total folate (ng/mL)
          LBDHDD, #	Serum HDL cholesterol (mg/dL)
          LBDLYMNO, #	Lymphocyte number (Coulter) (1000 cells/uL)
          LBDMONO, #	Monocyte number (1000 cells/uL)
          LBDNENO, #	Segmented neutrophils number (1000 cell/uL)
          LBDRFO, #	RBC folate (ng/mL)
          LBDSALSI, #	Serum albumin:  SI (g/L)
          LBDSBUSI, #	Serum Blood Urea Nitrogen: SI (mmol/L)
          LBDSCHSI, #	Serum total cholesterol:  SI (mmol/L)
          LBDSPHSI, #	Serum phosphorus: SI (mmol/L)
          LBDSTBSI, #	Serum total bilirubin:  SI (umol/L)
          LBDSTPSI, #	Serum total protein:  SI (g/L)
          LBDSUASI, #	Serum uric acid:  SI (umol/L)
          LBXBAPCT, #	Basophils percent (%)
          LBXEOPCT, #	Eosinophils percent (%)
          LBXGH, #	Glycohemoglobin (%)
          LBXHA, #	Serum hepatitis A antibody (Anti-HAV)
          LBXHBC, #	Serum hepatitis B core antibody
          LBXHBS, #	Serum hepatitis B surface antibody (anti-HBs)
          LBXHGB, #	Hemoglobin (g/dL)
          LBXLYPCT, #	Lymphocyte percent (Coulter) (%)
          LBXMC, #	Mean cell hemoglobin concentration (g/dL)
          LBXMCHSI, #	Mean cell hemoglobin (pg)
          LBXMCVSI, #	Mean cell volume (fL)
          LBXMOPCT, #	Monocyte percent (%)
          LBXMPSI, #	Mean platelet volume:  SI (fL)
          LBXNEPCT, #	Segmented neutrophils percent (%)
          LBXPLTSI, #	Platelet count: SI (1000 cells/uL)
          LBXRBCSI, #	Red blood cell count: SI (million cells/uL)
          LBXRDW, #	Red cell distribution width (%)
          LBXSAPSI, #	Serum alkaline phosphatase:   SI (U/L)
          LBXSASSI, #	Aspartate aminotransferase AST: SI (U/L)
          LBXSATSI, #	Alanine aminotransferase ALT: SI (U/L)
          LBXSBU, #	Serum blood urea nitrogen (mg/dL)
          LBXSC3SI, #	Serum bicarbonate:  SI (mmol/L)
          LBXSCR, #	Serum creatinine (mg/dL)
          LBXSGL, #	Serum glucose (mg/dL)
          LBXSLDSI, #	Serum lactate dehydrogenase: SI (U/L)
          LBXSTB, #	Serum total bilirubin (mg/dL)
          LBXSTP, #	Serum total protein (g/dL)
          LBXSUA, #	Serum uric acid (mg/dL)
          LBXWBCSI, #	White blood cell count: SI (1000 cells/uL)
          VNAVEBPXSY, #	Systolic: Average blood pressure (mm Hg)
          VNEGFR, #	Estimated Glomerular Filtration Rate (mL/min/1.73 m2)
          VNEGFRADJ, #	Estimated Glomerular Filtration Rate race-adjusted (mL/min/1.73 m2)
          VNHEMACRIT, #	Hematocrit (%)
          VNTOTHDRATIO) #	Ratio of Total to HDL Cholesterol (-)

### Causative Only
NHANES_RESPONSE_CAUS <- response_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
          BMXWT, #	Weight (kg)
          LBDFOT, #	Serum total folate (ng/mL) - Nutrition
          LBDHDD, #	Serum HDL cholesterol (mg/dL) - Metabolism and Lipid Markers, lifestyle
          LBDRFO, #	RBC folate (ng/mL) - Nutrition
          LBDSALSI, #	Serum albumin:  SI (g/L) - Nutrition, kidney or liver issues
          LBDSPHSI, #	Serum phosphorus: SI (mmol/L) - Kidney damage indicator and nutrition or thyroid issues
          LBDSUASI, #	Serum uric acid:  SI (umol/L) - Kidneyy/liver/thyroid problems, alcohol consumption, Obesity, Diabetes, cancers
          LBXGH, #	Glycohemoglobin (%) - Diabetes
          LBXSAPSI, #	Serum alkaline phosphatase:   SI (U/L) - kidney/liver disease, cancers, cirhhossis
          LBXSASSI, #	Aspartate aminotransferase AST: SI (U/L) - Liver Function, alcohol abuse, cirrhosis,
          LBXSATSI, #	Alanine aminotransferase ALT: SI (U/L)  - Liver Function
          LBXSBU, #	Serum blood urea nitrogen (mg/dL) - Metabolism, kidney/liver function, nutrition
          LBXSCR, #	Serum creatinine (mg/dL) - Renal Function
          LBXSGL, #	Serum glucose (mg/dL) - Matabolism, kidney function, diet, physical activity, nutrition
          LBXSLDSI, #	Serum lactate dehydrogenase: SI (U/L) - liver function, cancers
          LBXSTB, #	Serum total bilirubin (mg/dL)  - Liver Function
          LBXSUA, #	Serum uric acid (mg/dL) - Matabolism, kidney function, nutrition
          VNTOTHDRATIO) #	Ratio of Total to HDL Cholesterol (-)



```
:::

::: {style="margin-bottom:50px;"}
Other Chemicals data

```{r}
NHANES_SMOKE <- chemicals_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         LBXCOT) #	Serum cotinine (ng/mL) - Recent nicotine exposure from tobacco smoke

names(NHANES_SMOKE)[4] <- "COTININE"

```
:::

::: {style="margin-bottom:50px;"}
## Clean Race Data {style="margin-bottom:50px;"}

```{r}
NHANES_RACE <- demographics_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         RIDRETH1) # Recoded race category

# Rename variable
names(NHANES_RACE)[4] <- "RACE"

# This code appears unnecessary
NHANES_RACE <- NHANES_RACE %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         RACE)

# Change race to categorical factor
NHANES_RACE$RACE <- factor(NHANES_RACE$RACE)
```
:::

::: {style="margin-bottom:50px;"}
## Clean Education Data

```{r}
NHANES_EDUCATION <- demographics_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         DMDEDUC, # (SP Interview Version) What is the highest grade or level of school {you have/SP has} completed or the highest degree {you have/s/he has} received?
         DMDEDUC2,# Education level - Adults 20+
         DMDEDUC3) # Education level - Children/Youth 6-19


## Change the variables to ensure they are measured in the same scale
NHANES_EDUCATION$DMDEDUC2 <- ifelse(NHANES_EDUCATION$DMDEDUC2 == 1 | NHANES_EDUCATION$DMDEDUC2 == 2, 
                                    1,
                                    ifelse(NHANES_EDUCATION$DMDEDUC2 ==  3, 2,
                                           ifelse(NHANES_EDUCATION$DMDEDUC2 == 4 | NHANES_EDUCATION$DMDEDUC2 == 5,
                                                  3,
                                                  ifelse(NHANES_EDUCATION$DMDEDUC2 == 7,
                                                         7,
                                                         ifelse(NHANES_EDUCATION$DMDEDUC2 == 9,
                                                                9,
                                                                NA)))))

NHANES_EDUCATION$DMDEDUC3 <- ifelse(NHANES_EDUCATION$DMDEDUC3 <= 12 | NHANES_EDUCATION$DMDEDUC3 == 55 | NHANES_EDUCATION$DMDEDUC3 == 66, 1,
                                    ifelse(NHANES_EDUCATION$DMDEDUC3 == 13 | NHANES_EDUCATION$DMDEDUC3 == 14, 2,
                                           ifelse(NHANES_EDUCATION$DMDEDUC3 == 15, 3,
                                                  ifelse(NHANES_EDUCATION$DMDEDUC3 == 77, 7,
                                                         ifelse(NHANES_EDUCATION$DMDEDUC3 == 99, 9, NA)))))


NHANES_EDUCATION$EDUCATION <- ifelse(!is.na(NHANES_EDUCATION$DMDEDUC2),
                                     NHANES_EDUCATION$DMDEDUC2,
                                     ifelse(!is.na(NHANES_EDUCATION$DMDEDUC),
                                            NHANES_EDUCATION$DMDEDUC, 
                                            ifelse(!is.na(NHANES_EDUCATION$DMDEDUC3), NHANES_EDUCATION$DMDEDUC3, NA)))

NHANES_EDUCATION <- NHANES_EDUCATION %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         EDUCATION)
NHANES_EDUCATION$EDUCATION <- factor(NHANES_EDUCATION$EDUCATION)
```
:::

::: {style="margin-bottom:50px;"}
## Clean PFAS Data

```{r}
NHANES_PFAS <- chemicals_clean %>%
  select(
    SEQN, 
    SEQN_new, 
    SDDSRVYR,
    VNURXPFBA, 
    LBXPFBS, 
    VNURXPFBS, 
    VNURXPFHA, 
    LBXPFHS, 
    VNURXPFHS,
    VNURXGENX)

# Get rid of any rows that are missing data for all PFAS
NHANES_PFAS <- NHANES_PFAS %>%
  filter(if_any(c(VNURXPFBA, 
                  LBXPFBS, 
                  VNURXPFBS, 
                  VNURXPFHA, 
                  LBXPFHS, 
                  VNURXPFHS,
                  VNURXGENX), 
                ~ !is.na(.)))
```
:::

::: {style="margin-bottom:50px;"}
::: {style="margin-bottom:50px;"}
## Clean Cancer Data

```{r}
NHANES_HEALTH <- questionnaire_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR, 
         MCQ240M,	# Age when liver cancer was first diagnosed
         MCQ240BB,#	Age when thyroid cancer was first diagnosed
         MCQ240N,	# Age when lung cancer was first diagnosed
         MCQ240J, #	Age when kidney cancer was first diagnosed
         HAC3OS,  #	Where was cancer located when first told
         MCQ230A, #	1st cancer - what kind was it?
         MCQ230B, #	2nd cancer - what kind was it?
         MCQ230C,	# 3rd cancer - what kind was it?
         MCQ230D) #	More than 3 kinds of cancer - What kind of cancer was it?

## LIVER CANCER CLASSIFICATION
# Liver Cancer Class
NHANES_HEALTH$LIVER_CANCER_CLASS <- ifelse(
  (!is.na(NHANES_HEALTH$MCQ240M) & 
     NHANES_HEALTH$MCQ240M > 0 & 
     NHANES_HEALTH$MCQ240M < 100) |
    (!is.na(NHANES_HEALTH$MCQ230A) & NHANES_HEALTH$MCQ230A == 22) |
    (!is.na(NHANES_HEALTH$MCQ230B) & NHANES_HEALTH$MCQ230B == 22) |
    (!is.na(NHANES_HEALTH$MCQ230C) & NHANES_HEALTH$MCQ230C == 22) |
    (!is.na(NHANES_HEALTH$MCQ230D) & NHANES_HEALTH$MCQ230D == 22),
  1,
  0
)

# Thyroid Cancer Class
NHANES_HEALTH$THYROID_CANCER_CLASS <- ifelse(
  (!is.na(NHANES_HEALTH$MCQ240BB) & 
     NHANES_HEALTH$MCQ240BB > 0 & 
     NHANES_HEALTH$MCQ240BB < 100) |
    (!is.na(NHANES_HEALTH$MCQ230A) & NHANES_HEALTH$MCQ230A == 37) |
    (!is.na(NHANES_HEALTH$MCQ230B) & NHANES_HEALTH$MCQ230B == 37) |
    (!is.na(NHANES_HEALTH$MCQ230C) & NHANES_HEALTH$MCQ230C == 37) |
    (!is.na(NHANES_HEALTH$MCQ230D) & NHANES_HEALTH$MCQ230D == 37),
  1,
  0
)

# Lung Cancer Class
NHANES_HEALTH$LUNG_CANCER_CLASS <- ifelse(
  (!is.na(NHANES_HEALTH$MCQ240N) & 
     NHANES_HEALTH$MCQ240N > 0 & 
     NHANES_HEALTH$MCQ240N < 100) |
    (!is.na(NHANES_HEALTH$MCQ230A) & NHANES_HEALTH$MCQ230A == 23) |
    (!is.na(NHANES_HEALTH$MCQ230B) & NHANES_HEALTH$MCQ230B == 23) |
    (!is.na(NHANES_HEALTH$MCQ230C) & NHANES_HEALTH$MCQ230C == 23) |
    (!is.na(NHANES_HEALTH$MCQ230D) & NHANES_HEALTH$MCQ230D == 23),
  1,
  0
)

# Kidney Cancer Class
NHANES_HEALTH$KIDNEY_CANCER_CLASS <- ifelse(
  (!is.na(NHANES_HEALTH$MCQ240J) & 
     NHANES_HEALTH$MCQ240J > 0 &
     NHANES_HEALTH$MCQ240J < 100) |
    (!is.na(NHANES_HEALTH$MCQ230A) & NHANES_HEALTH$MCQ230A == 19) |
    (!is.na(NHANES_HEALTH$MCQ230B) & NHANES_HEALTH$MCQ230B == 19) |
    (!is.na(NHANES_HEALTH$MCQ230C) & NHANES_HEALTH$MCQ230C == 19) |
    (!is.na(NHANES_HEALTH$MCQ230D) & NHANES_HEALTH$MCQ230D == 19),
  1,
  0
)

# Convert cancer classification to a factor:
NHANES_HEALTH$LIVER_CANCER_CLASS <- factor(NHANES_HEALTH$LIVER_CANCER_CLASS)
NHANES_HEALTH$THYROID_CANCER_CLASS <- factor(NHANES_HEALTH$THYROID_CANCER_CLASS)
NHANES_HEALTH$LUNG_CANCER_CLASS <- factor(NHANES_HEALTH$LUNG_CANCER_CLASS)
NHANES_HEALTH$KIDNEY_CANCER_CLASS <- factor(NHANES_HEALTH$KIDNEY_CANCER_CLASS)

# View total cancer diagnosis:
sum(NHANES_HEALTH$LIVER_CANCER_CLASS == 1)
sum(NHANES_HEALTH$THYROID_CANCER_CLASS == 1)
sum(NHANES_HEALTH$LUNG_CANCER_CLASS == 1)
sum(NHANES_HEALTH$KIDNEY_CANCER_CLASS == 1)

# Subset the data to include only cancer classification:
NHANES_HEALTH <- NHANES_HEALTH %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         LIVER_CANCER_CLASS,
         THYROID_CANCER_CLASS,
         LUNG_CANCER_CLASS,
         KIDNEY_CANCER_CLASS)
```
:::
:::

::: {style="margin-bottom:50px;"}
## Join the Data {style="margin-bottom:50px;"}

```{r}

nhanes_merged <- left_join(NHANES_PFAS, 
                           NHANES_HEALTH, 
                           by = c("SEQN",
                                  "SEQN_new", 
                                  "SDDSRVYR")) %>%

  left_join(., 
             NHANES_SMOKE, 
             by = c("SEQN",
                    "SEQN_new", 
                    "SDDSRVYR")) %>%
  left_join(., 
            NHANES_AGE, 
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(., 
            NHANES_GENDER, 
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_BMI,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_EDUCATION,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_eGFR,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_RACE,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_RESPONSE_CAUS,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR"))%>%
left_join(.,
            NHANES_INCOME,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR"))
```
:::

::: {style="margin-bottom:50px;"}
## Compile PFAS Separately

```{r}
compiled.PFHxS.SERUM <- nhanes_merged %>%
  select(-c(VNURXPFBA, 
            VNURXPFHS,
            LBXPFBS, 
            VNURXPFBS, 
            VNURXPFHA,
            VNURXGENX)) 
compiled.PFHxS.SERUM <- na.omit(compiled.PFHxS.SERUM)

# Remove identifiers 
compiled.PFHxS.SERUM <- compiled.PFHxS.SERUM %>%
  select(-c(SEQN,SEQN_new, SDDSRVYR))

# Rename response column and change to numeric
names(compiled.PFHxS.SERUM)[1] = "Serum_PFHxS"
compiled.PFHxS.SERUM$Serum_PFHxS <- as.numeric(compiled.PFHxS.SERUM$Serum_PFHxS)

```
:::

::: {style="margin-bottom:50px;"}
## Export the file to your working directory

```{r}

write.csv(compiled.PFHxS.SERUM, "Compiled PFHxS Serum.csv")
```
:::
