---
title: "Untitled"
author: "Nolan Moody (UCID 30044615)"
date: "2024-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

::: {style="margin-bottom:50px;"}
## Load Packages, Get Data

```{r}
install.packages("tidyverse") # Needed to pipe functions together
library("tidyverse") 
install.packages("mosaic")
library(mosaic)
install.packages("ggplot2")
library(ggplot2)
setwd("C:/Users/nolan/OneDrive/Desktop/DATA501 Project Files") 
load("./w - nhanes_1988_2018.RData")
```
:::

::: {style="margin-bottom:50px;"}
### Clean Age Data

```{r}
NHANES_AGE <- demographics_clean %>%
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier Data
         RIDAGEYR) # Recoded and cleaned age of all participants

# Change column name for easier processing
names(NHANES_AGE)[4] = "AGE"

# Convert age to numeric
NHANES_AGE$AGE <- as.numeric(NHANES_AGE$AGE)
```
:::

::: {style="margin-bottom:50px;"}
## Clean Gender Data

```{r}
NHANES_GENDER <- demographics_clean %>%
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier Data
         RIAGENDR) 	 # Recoded and cleaned gender for all participants

# Rename column for easier processing.
names(NHANES_GENDER)[4] <- "GENDER"

# Change categories. Male is originally measured as 1, female as 2. Change male to 0 and female to 1 to help with regression design.
NHANES_GENDER$GENDER <- ifelse(NHANES_GENDER$GENDER == 1, 0, 1)

# Change gender to categorical factor.
NHANES_GENDER$GENDER <- factor(NHANES_GENDER$GENDER)
```
:::

::: {style="margin-bottom:50px;"}
## Clean BMI Data

```{r}
NHANES_BMI_2 <- response_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         BMXBMI, # Adult BMI
         BMDBMIC) # Youth category	
# Categories for youth BMI:
# 1 = Underweight
# 2 = Normal Weight
# 3 = Overweight
# 4 = Obese

# Create an alternate measure of BMI that represents categories:
NHANES_BMI_2$BMI_CAT <- 
  ifelse(!is.na(NHANES_BMI_2$BMDBMIC), NHANES_BMI_2$BMDBMIC,
     ifelse(is.na(NHANES_BMI_2$BMXBMI), NA, 
            ifelse(NHANES_BMI_2$BMXBMI < 18.5, 1,
                    ifelse(NHANES_BMI_2$BMXBMI < 25, 2,
                           ifelse(NHANES_BMI_2$BMXBMI < 30, 3, 4)))))
# Change column name for convenience:
names(NHANES_BMI_2)[4] = "BMI"

# Subset to keep only data of interest for the primary BMI df:
NHANES_BMI <- NHANES_BMI_2 %>% 
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier variables
         BMI) # BMI as a number
# Ensure BMI is numeric
NHANES_BMI$BMI <- as.numeric(NHANES_BMI$BMI)

NHANES_BMI_alt <- NHANES_BMI_2 %>% 
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier variables
         BMI_CAT) # BMI category

```
:::

::: {style="margin-bottom:50px;"}
## Clean eGFR Data {style="margin-bottom:50px;"}

```{r}
NHANES_eGFR  <- response_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         VNEGFR) # Measured eGFR

# Rename variable for convenience
names(NHANES_eGFR)[4] = "eGFR"

# Ensure numerical measure
NHANES_eGFR$eGFR <- as.numeric(NHANES_eGFR$eGFR)
```
:::

::: {style="margin-bottom:50px;"}
## **Other Demographic Variables of Interest** {style="margin-bottom:50px;"}

INDFMIN2 Total family income (reported as a range value in dollars) 132610

INDFMPIR Poverty income ratio (PIR) - a ratio of family income to poverty threshold 122704

```{r}
NHANES_INCOME <- demographics_clean %>%
  select(SEQN, SEQN_new, SDDSRVYR, # Identifier Data
         INDFMIN2, # Total Family Income
         INDFMPIR) # Poverty ratio
NHANES_INCOME$INDFMIN2 <- as.factor(NHANES_INCOME$INDFMIN2)
names(NHANES_INCOME)[4] = "INCOME_CAT"
names(NHANES_INCOME)[5] = "POVERTY_RATIO"
# Important to note: These two variables are highly correlated at ~0.85
```
:::

::: {style="margin-bottom:50px;"}
## Other Response Variables

```{r}
### Correlative and Causative
NHANES_RESPONSE_CORR <- response_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
          BMXWT, #	Weight (kg)
          BPXDI1, #	Diastolic: Blood pressure (first reading) mm Hg
          BPXML1, #	MIL: maximum inflation levels (mm Hg)
          BPXPLS, #	60 sec. pulse (30 sec. pulse * 2)
          LBDBANO, #	Basophils number (1000 cells/uL)
          LBDEONO, #	Eosinophils number (1000 cells/uL)
          LBDFOT, #	Serum total folate (ng/mL)
          LBDHDD, #	Serum HDL cholesterol (mg/dL)
          LBDLYMNO, #	Lymphocyte number (Coulter) (1000 cells/uL)
          LBDMONO, #	Monocyte number (1000 cells/uL)
          LBDNENO, #	Segmented neutrophils number (1000 cell/uL)
          LBDRFO, #	RBC folate (ng/mL)
          LBDSALSI, #	Serum albumin:  SI (g/L)
          LBDSBUSI, #	Serum Blood Urea Nitrogen: SI (mmol/L)
          LBDSCHSI, #	Serum total cholesterol:  SI (mmol/L)
          LBDSPHSI, #	Serum phosphorus: SI (mmol/L)
          LBDSTBSI, #	Serum total bilirubin:  SI (umol/L)
          LBDSTPSI, #	Serum total protein:  SI (g/L)
          LBDSUASI, #	Serum uric acid:  SI (umol/L)
          LBXBAPCT, #	Basophils percent (%)
          LBXEOPCT, #	Eosinophils percent (%)
          LBXGH, #	Glycohemoglobin (%)
          LBXHA, #	Serum hepatitis A antibody (Anti-HAV)
          LBXHBC, #	Serum hepatitis B core antibody
          LBXHBS, #	Serum hepatitis B surface antibody (anti-HBs)
          LBXHGB, #	Hemoglobin (g/dL)
          LBXLYPCT, #	Lymphocyte percent (Coulter) (%)
          LBXMC, #	Mean cell hemoglobin concentration (g/dL)
          LBXMCHSI, #	Mean cell hemoglobin (pg)
          LBXMCVSI, #	Mean cell volume (fL)
          LBXMOPCT, #	Monocyte percent (%)
          LBXMPSI, #	Mean platelet volume:  SI (fL)
          LBXNEPCT, #	Segmented neutrophils percent (%)
          LBXPLTSI, #	Platelet count: SI (1000 cells/uL)
          LBXRBCSI, #	Red blood cell count: SI (million cells/uL)
          LBXRDW, #	Red cell distribution width (%)
          LBXSAPSI, #	Serum alkaline phosphatase:   SI (U/L)
          LBXSASSI, #	Aspartate aminotransferase AST: SI (U/L)
          LBXSATSI, #	Alanine aminotransferase ALT: SI (U/L)
          LBXSBU, #	Serum blood urea nitrogen (mg/dL)
          LBXSC3SI, #	Serum bicarbonate:  SI (mmol/L)
          LBXSCR, #	Serum creatinine (mg/dL)
          LBXSGL, #	Serum glucose (mg/dL)
          LBXSLDSI, #	Serum lactate dehydrogenase: SI (U/L)
          LBXSTB, #	Serum total bilirubin (mg/dL)
          LBXSTP, #	Serum total protein (g/dL)
          LBXSUA, #	Serum uric acid (mg/dL)
          LBXWBCSI, #	White blood cell count: SI (1000 cells/uL)
          VNAVEBPXSY, #	Systolic: Average blood pressure (mm Hg)
          VNEGFR, #	Estimated Glomerular Filtration Rate (mL/min/1.73 m2)
          VNEGFRADJ, #	Estimated Glomerular Filtration Rate race-adjusted (mL/min/1.73 m2)
          VNHEMACRIT, #	Hematocrit (%)
          VNTOTHDRATIO) #	Ratio of Total to HDL Cholesterol (-)

### Causative Only
NHANES_RESPONSE_CAUS <- response_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
          BMXWT, #	Weight (kg)
          LBDFOT, #	Serum total folate (ng/mL) - Nutrition
          LBDHDD, #	Serum HDL cholesterol (mg/dL) - Metabolism and Lipid Markers, lifestyle
          LBDRFO, #	RBC folate (ng/mL) - Nutrition
          LBDSALSI, #	Serum albumin:  SI (g/L) - Nutrition, kidney or liver issues
          LBDSPHSI, #	Serum phosphorus: SI (mmol/L) - Kidney damage indicator and nutrition or thyroid issues
          LBDSUASI, #	Serum uric acid:  SI (umol/L) - Kidney/liver/thyroid problems, alcohol consumption, Obesity, Diabetes, cancers
          LBXGH, #	Glycohemoglobin (%) - Diabetes
          LBXSAPSI, #	Serum alkaline phosphatase:   SI (U/L) - kidney/liver disease, cancers, cirhhossis
          LBXSASSI, #	Aspartate aminotransferase AST: SI (U/L) - Liver Function, alcohol abuse, cirrhosis,
          LBXSATSI, #	Alanine aminotransferase ALT: SI (U/L)  - Liver Function
          LBXSBU, #	Serum blood urea nitrogen (mg/dL) - Metabolism, kidney/liver function, nutrition
          LBXSCR, #	Serum creatinine (mg/dL) - Renal Function
          LBXSGL, #	Serum glucose (mg/dL) - Matabolism, kidney function, diet, physical activity, nutrition
          LBXSLDSI, #	Serum lactate dehydrogenase: SI (U/L) - liver function, cancers
          LBXSTB, #	Serum total bilirubin (mg/dL)  - Liver Function
          LBXSUA, #	Serum uric acid (mg/dL) - Matabolism, kidney function, nutrition
          VNTOTHDRATIO) #	Ratio of Total to HDL Cholesterol (-)



```
:::

::: {style="margin-bottom:50px;"}
Other Chemicals data

```{r}
NHANES_SMOKE <- chemicals_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         LBXCOT) #	Serum cotinine (ng/mL) - Recent nicotine exposure from tobacco smoke

names(NHANES_SMOKE)[4] <- "COTININE"

```
:::

::: {style="margin-bottom:50px;"}
## Clean Race Data {style="margin-bottom:50px;"}

```{r}
NHANES_RACE <- demographics_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         RIDRETH1) # Recoded race category

# Rename variable
names(NHANES_RACE)[4] <- "RACE"

# This code appears unnecessary
NHANES_RACE <- NHANES_RACE %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         RACE)

# Change race to categorical factor
NHANES_RACE$RACE <- factor(NHANES_RACE$RACE)
```
:::

::: {style="margin-bottom:50px;"}
## Clean Education Data

```{r}
NHANES_EDUCATION <- demographics_clean %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         DMDEDUC, # (SP Interview Version) What is the highest grade or level of school {you have/SP has} completed or the highest degree {you have/s/he has} received?
         DMDEDUC2,# Education level - Adults 20+
         DMDEDUC3) # Education level - Children/Youth 6-19


## Change the variables to ensure they are measured in the same scale
NHANES_EDUCATION$DMDEDUC2 <- ifelse(NHANES_EDUCATION$DMDEDUC2 == 1 | NHANES_EDUCATION$DMDEDUC2 == 2, 
                                    1,
                                    ifelse(NHANES_EDUCATION$DMDEDUC2 ==  3, 2,
                                           ifelse(NHANES_EDUCATION$DMDEDUC2 == 4 | NHANES_EDUCATION$DMDEDUC2 == 5,
                                                  3,
                                                  ifelse(NHANES_EDUCATION$DMDEDUC2 == 7,
                                                         7,
                                                         ifelse(NHANES_EDUCATION$DMDEDUC2 == 9,
                                                                9,
                                                                NA)))))

NHANES_EDUCATION$DMDEDUC3 <- ifelse(NHANES_EDUCATION$DMDEDUC3 <= 12 | NHANES_EDUCATION$DMDEDUC3 == 55 | NHANES_EDUCATION$DMDEDUC3 == 66, 1,
                                    ifelse(NHANES_EDUCATION$DMDEDUC3 == 13 | NHANES_EDUCATION$DMDEDUC3 == 14, 2,
                                           ifelse(NHANES_EDUCATION$DMDEDUC3 == 15, 3,
                                                  ifelse(NHANES_EDUCATION$DMDEDUC3 == 77, 7,
                                                         ifelse(NHANES_EDUCATION$DMDEDUC3 == 99, 9, NA)))))


NHANES_EDUCATION$EDUCATION <- ifelse(!is.na(NHANES_EDUCATION$DMDEDUC2),
                                     NHANES_EDUCATION$DMDEDUC2,
                                     ifelse(!is.na(NHANES_EDUCATION$DMDEDUC),
                                            NHANES_EDUCATION$DMDEDUC, 
                                            ifelse(!is.na(NHANES_EDUCATION$DMDEDUC3), NHANES_EDUCATION$DMDEDUC3, NA)))

NHANES_EDUCATION <- NHANES_EDUCATION %>%
  select(SEQN, 
         SEQN_new, 
         SDDSRVYR,
         EDUCATION)
NHANES_EDUCATION$EDUCATION <- factor(NHANES_EDUCATION$EDUCATION)
```
:::

::: {style="margin-bottom:50px;"}
## Clean PFAS Data

```{r}
NHANES_PFAS <- chemicals_clean %>%
  select(
    SEQN, 
    SEQN_new, 
    SDDSRVYR,
    VNURXPFBA, 
    LBXPFBS, 
    VNURXPFBS, 
    VNURXPFHA, 
    LBXPFHS, 
    VNURXPFHS,
    VNURXGENX)

# Get rid of any rows that are missing data for all PFAS
NHANES_PFAS <- NHANES_PFAS %>%
  filter(if_any(c(VNURXPFBA, 
                  LBXPFBS, 
                  VNURXPFBS, 
                  VNURXPFHA, 
                  LBXPFHS, 
                  VNURXPFHS,
                  VNURXGENX), 
                ~ !is.na(.)))
```
:::

::: {style="margin-bottom:50px;"}
## Join the Data {style="margin-bottom:50px;"}

```{r}
nhanes_merged <- left_join(NHANES_PFAS, 
                           NHANES_SMOKE, 
                           by = c("SEQN",
                                  "SEQN_new", 
                                  "SDDSRVYR")) %>%
  left_join(., 
            NHANES_AGE, 
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(., 
            NHANES_GENDER, 
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_BMI,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_EDUCATION,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_eGFR,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_RACE,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR")) %>%
  left_join(.,
            NHANES_RESPONSE_CAUS,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR"))%>%
left_join(.,
            NHANES_INCOME,
            by = c("SEQN",
                   "SEQN_new", 
                   "SDDSRVYR"))
```
:::

::: {style="margin-bottom:50px;"}
## Compile PFAS Separately

```{r}
compiled.PFHxS.SERUM <- nhanes_merged %>%
  select(-c(VNURXPFBA, 
            VNURXPFHS,
            LBXPFBS, 
            VNURXPFBS, 
            VNURXPFHA,
            VNURXGENX)) 
compiled.PFHxS.SERUM <- na.omit(compiled.PFHxS.SERUM)

# Remove identifiers 
compiled.PFHxS.SERUM <- compiled.PFHxS.SERUM %>%
  select(-c(SEQN,SEQN_new, SDDSRVYR))

# Rename response column and change to numeric
names(compiled.PFHxS.SERUM)[1] = "Serum_PFHxS"
compiled.PFHxS.SERUM$Serum_PFHxS <- as.numeric(compiled.PFHxS.SERUM$Serum_PFHxS)

```
:::

::: {style="margin-bottom:50px;"}
## Apply GBM to identify important variables -\> Both Genders

```{r}
install.packages("gbm")
install.packages("xgboost")
install.packages("mltools")
install.packages("randomForest")
library(xgboost)
library(mltools)
library(data.table)
```

```{r}

######### XGBOOST ALGORITHM ##################


hot.PFHxS <- as.matrix(one_hot(as.data.table(compiled.PFHxS.SERUM)))

set.seed(1)
train = sample(nrow(hot.PFHxS), floor(0.8 * nrow(hot.PFHxS)), replace = F)

xgb.PFHxS <- xgboost(data = hot.PFHxS[train, -1], # evrything but label variable
               label = hot.PFHxS[train, 1],
               nrounds = 10000, # like n trees
               nfold = 5,# 5 fold cv
               objective = "reg:squarederror", # regrssion validation
               eta = 0.1, # learning rate
               verbose = FALSE,
               early_stopping_rounds = 20)

xgb.pred <- predict(xgb.PFHxS, hot.PFHxS[-train, -1])
sqrt(mean((xgb.pred - hot.PFHxS[-train, 1])^2))
# Returns a RMSE of 0.002909

importance_matrix <- xgb.importance(model = xgb.PFHxS)


xgb.plot.importance(importance_matrix, top_n = 30, cex = 0.35)
# Important Variables:
# LBDFRO
# AGE
# COTININE
# eGFR
# BMI
# POVERTY_RATIO
# LBXSTB
# VNTOTHDRATIO
# LBXSAPSI
# LBDFOT
# LBXSGL
# LBDHDD
# LBDSPHSI
# LBDSALSI
# LBDSUASI

############# GBM ALGORITHM ##########

library(gbm)
set.seed(1)

gbm2 <- gbm(Serum_PFHxS ~ ., distribution = "gaussian",
             data = compiled.PFHxS.SERUM[train, ], n.trees = 1000,
             interaction.depth = 1, shrinkage = 0.01,
            cv.folds = 5, n.cores = NULL, verbose = F)
gbm.pred <- predict(gbm2, compiled.PFHxS.SERUM[-train, ])
sqrt(mean((gbm.pred - compiled.PFHxS.SERUM$Serum_PFHxS[-train])^2))
# Returns RMSE of 2.427

summary(gbm2, las = 2)
View(as.data.frame(summary(gbm2, las = 2)[1]))

# Important Variables:
# LBDRFO	LBDRFO	16.3981082		
# LBXSTB	LBXSTB	13.5616029		
# GENDER	GENDER	13.4603037		
# AGE	AGE	13.0106397		
# RACE	RACE	8.3897153		
# LBDSALSI	LBDSALSI	6.5230413		
# INCOME_CAT	INCOME_CAT	5.4265507		
# LBXSCR	LBXSCR	4.1071891		
# LBXSAPSI	LBXSAPSI	3.8757364		
# LBXSLDSI	LBXSLDSI	3.4800454	
# POVERTY_RATIO	POVERTY_RATIO	3.3932959		
# COTININE	COTININE	2.5756765		
# BMXWT	BMXWT	1.6880387		
# LBXSASSI	LBXSASSI	1.2705953		
# LBXSGL	LBXSGL	1.0298356	

############## Random Forest ##########

library(randomForest)
set.seed(1)
rf <- randomForest(Serum_PFHxS ~., data = compiled.PFHxS.SERUM[train, ], ntree = 100, importance = T)

rf.pred <- predict(rf, newdata = compiled.PFHxS.SERUM[-train,])
sqrt(mean((rf.pred - compiled.PFHxS.SERUM$Serum_PFHxS[-train])^2))
# Returns RMSE of 1.137

plot(rf)
View(as.data.frame(importance(rf)))
varImpPlot(rf, n.var = 27)


# Important Variables (MSE):
# AGE
# LBDFOT
# GENDER
# LBDSUASI
# LBXSCR
# LBXSUA
# LBDSPHSI
# LBXSATSI
# LBXSAPSI
# RACE
# LBDRFO
# LBXSBU
# eGFR
# POVERTY_RATIO
# LBDHDD

```

```{r}
# Top 10 Variables:
# AGE	
# LBDRFO	RBC folate (ng/mL)
# LBDFOT Serum total folate (ng/mL)
# LBXSAPSI	Serum alkaline phosphatase:   SI (U/L)
# eGFR
# POVERTY_RATIO	
# COTININE	
# LBDSPHSI	Serum phosphorus: SI (mmol/L)
# LBXSCR	Serum creatinine (mg/dL)
# LBDSUASI	Serum uric acid:  SI (umol/L)

```
:::

::: {style="margin-bottom:50px;"}
## GBM for Importance - Male Only

```{r}

male.PFHxS = filter(compiled.PFHxS.SERUM, GENDER == 0)

male.PFHxS <- male.PFHxS %>%
  select(-GENDER)

hot.PFHxS.male <- as.matrix(one_hot(as.data.table(male.PFHxS)))

set.seed(1)
train.male = sample(nrow(male.PFHxS), floor(0.8 * nrow(male.PFHxS)), replace = F)

xgb.PFHxS.male <- xgboost(data = hot.PFHxS.male[train.male, -1], # evrything but label variable
               label = hot.PFHxS.male[train.male, 1],
               nrounds = 10000, # like n trees
               nfold = 5,# 5 fold cv
               objective = "reg:squarederror", # regrssion validation
               eta = 0.1, # learning rate
               verbose = FALSE,
               early_stopping_rounds = 20)

xgb.pred.male <- predict(xgb.PFHxS.male, hot.PFHxS.male[-train.male, -1])
sqrt(mean((xgb.pred.male - hot.PFHxS.male[-train.male, 1])^2))
# Returns a RMSE of 3.9987

importance_matrix_male <- xgb.importance(model = xgb.PFHxS.male)


xgb.plot.importance(importance_matrix_male, top_n = 30, cex = 0.35)


#### Important Variables
# Cotinine
# LBDRFO
# LBXSTB
# BMI
# LBDFOT
# eGFR
# POVERTY_RATIO
# VNTOTHDRATIO
# AGE
# LBXSAPSI
# LBXSGl
# LBDSUASI
# LBDHDD
# LBXSLDSI
# BMXWT


############# GBM ALGORITHM ##########

set.seed(1)

gbm2.male <- gbm(Serum_PFHxS ~ ., distribution = "gaussian",
             data = male.PFHxS[train.male,], n.trees = 1000,
             interaction.depth = 1, shrinkage = 0.01,
            cv.folds = 5, n.cores = NULL, verbose = F)
gbm.pred.male <- predict(gbm2.male, male.PFHxS[-train.male, ])
sqrt(mean((gbm.pred.male - male.PFHxS$Serum_PFHxS[-train.male])^2))
# Returns a RMSE of 3.88

summary(gbm2.male, las = 2)
View(as.data.frame(summary(gbm2.male, las = 2)[1]))
### Important Variables
# INCOME_CAT	INCOME_CAT	12.63865215		
# RACE	RACE	12.23707425		
# LBDRFO	LBDRFO	11.31905535		
# LBXSTB	LBXSTB	10.97815003		
# LBXSAPSI	LBXSAPSI	8.21074281		
# AGE	AGE	5.54771191		
# LBXSGL	LBXSGL	5.22341888		
# POVERTY_RATIO	POVERTY_RATIO	4.86233863		
# LBDSALSI	LBDSALSI	4.48721219		
# LBXSCR	LBXSCR	3.71276501	
# LBDSUASI	LBDSUASI	3.13748938		
# LBXSLDSI	LBXSLDSI	3.12665964		
# LBDFOT	LBDFOT	2.17836097		
# LBDSPHSI	LBDSPHSI	2.17160832		
# BMI	BMI	2.17130528	

############## Random Forest ##########
set.seed(1)
rf.male <- randomForest(Serum_PFHxS ~., data = male.PFHxS[train.male, ], ntree = 100, importance = T)

rf.pred.male <- predict(rf.male, newdata = male.PFHxS[-train.male,])
sqrt(mean((rf.pred.male - male.PFHxS$Serum_PFHxS[-train.male])^2))
# Returns a RMSE of 3.9

plot(rf.male)
importance(rf.male)
varImpPlot(rf.male)
varImpPlot(rf.male, n.var = 27)
View(as.data.frame(importance(rf.male)))
### Important Variables
# AGE
# BMXWT
# POVERTY_RATIO
# RACE
# VNTOTHDRATIO
# LBXSBU
# eGFR
# LBXSCR
# BMI
# LBXSAPSI
# LBDFOT
# LBDSUASI
# LBDSPHSI
# EDUCATION
# LBXSATSI

```

```{r}
# Top 10 Variables
# LBDRFO	RBC folate (ng/mL)
# AGE	
# BMI	
# LBXSAPSI	Serum alkaline phosphatase:   SI (U/L)
# RACE
# POVERTY_RATIO	
# VNTOTHDRATIO		Ratio of Total to HDL Cholesterol (-)
# LBXSGl		Serum glucose (mg/dL)
# eGFR
# LBDFOT	Serum total folate (ng/mL)

```
:::

::: {style="margin-bottom:50px;"}
GBM for Importance - Female Only

```{r}
female.PFHxS = filter(compiled.PFHxS.SERUM, GENDER == 1)

female.PFHxS <- female.PFHxS %>%
  select(-GENDER)

hot.PFHxS.female <- as.matrix(one_hot(as.data.table(female.PFHxS)))

set.seed(1)
train.female = sample(nrow(female.PFHxS), floor(0.8 * nrow(female.PFHxS)), replace = F)

xgb.PFHxS.female <- xgboost(data = hot.PFHxS.female[train.female, -1], # evrything but label variable
               label = hot.PFHxS.female[train.female, 1],
               nrounds = 10000, # like n trees 
               nfold = 5,# 5 fold cv
               objective = "reg:squarederror", # regrssion validation
               eta = 0.1, # learning rate
               verbose = FALSE,
               early_stopping_rounds = 20)

xgb.pred.female <- predict(xgb.PFHxS.female, hot.PFHxS.female[-train.female, -1])
sqrt(mean((xgb.pred.female - hot.PFHxS.female[-train.female, 1])^2))
# Returns a RMSE of 2.88

importance_matrix_female <- xgb.importance(model = xgb.PFHxS.female)

xgb.plot.importance(importance_matrix_female, top_n = 30, cex = 0.35)
xgb.plot.importance(importance_matrix_female, top_n = 15)

##### Important Variables
# AGE
# LBDRFO
# COTININE
# eGFR
# POVERTY_RATIO
# BMI
# LBDFOT
# VNTOTHDRATIO
# LBXSAPSI
# BMXWT
# LBXSTB
# LBXSGL
# LBDSPHSI
# LBXSLDSI
# LBDHDD


############# GBM ALGORITHM ##########

library(gbm)
set.seed(1)

gbm2.female <- gbm(Serum_PFHxS ~ ., distribution = "gaussian",
             data = female.PFHxS[train.female,], n.trees = 1000,
             interaction.depth = 1, shrinkage = 0.01,
            cv.folds = 5, n.cores = NULL, verbose = F)

gbm.pred.female <- predict(gbm2.female, female.PFHxS[-train.female, ])
sqrt(mean((gbm.pred.female - female.PFHxS$Serum_PFHxS[-train.female])^2))
# Returns a RMSE of 2.748

summary(gbm2.female, las = 2)
View(as.data.frame(summary(gbm2.female, las = 2)[1]))

##### Important Variables
# AGE	AGE	34.28830856		
# LBDRFO	LBDRFO	21.68539146		
# RACE	RACE	9.04015809		
# LBXSTB	LBXSTB	6.47010625		
# INCOME_CAT	INCOME_CAT	6.08562378		
# BMI	BMI	4.11706942		
# LBXSAPSI	LBXSAPSI	4.01857718		
# POVERTY_RATIO	POVERTY_RATIO	3.91994320		
# LBXSATSI	LBXSATSI	2.22078313		
# LBDSALSI	LBDSALSI	1.53813647	
# LBXSASSI	LBXSASSI	1.29028644		
# COTININE	COTININE	1.09530032		
# LBXSGL	LBXSGL	0.82939984		
# LBXGH	LBXGH	0.63933737		
# BMXWT	BMXWT	0.61907843

############## Random Forest ##########

set.seed(1)
library(randomForest)
rf.female <- randomForest(Serum_PFHxS ~., data = female.PFHxS[train.female, ], ntree = 100, importance = T)

rf.pred.female <- predict(rf.female, newdata = female.PFHxS[-train.female,])
sqrt(mean((rf.pred.female - female.PFHxS$Serum_PFHxS[-train.female])^2))
# Returns a RMSE of 2.77798

plot(rf.female)
importance(rf.female)
varImpPlot(rf.female, n.var = 15)
View(as.data.frame(importance(rf.female)))
##### Important Variables
# AGE
# LBDFOT
# LBDRFO
# POVERTY_RATIO
# COTININE
# LBXSUA
# LBXSBU
# EGFR
# LBXGH
# LBXSCR
# INCOME_CAT
# RACE
# BMXWT
# LBDSUASI
# LBXSASSI
```

```{r}
# Top 10 variables
# AGE	
# LBDRFO		RBC folate (ng/mL)
# BMI	
# POVERTY_RATIO	
# COTININE	
# BMXWT	Weight (kg)
# eGFR	
# LBXSAPSI		Serum alkaline phosphatase:   SI (U/L)
# RACE	
# LBXSTB	Serum total bilirubin (mg/dL)

```
:::

::: {style="margin-bottom:50px;"}
## Plot Important Variables with PFHxS

Variables identified as important:

AGE

BMI

URXUCR: Urinary creatinine (mg/dL)

eGFR

LBXRDW: Red cell distribution width (%)

BMXHT: Standing Height (cm)

URXUMA: Urinary albumin (ug/mL)

LBXLYPCT: Lymphocyte percent (Coulter) (%)

LBDRFO: RBC folate (ng/mL)

LBDFOT: Serum total folate (ng/mL)

```{r}

ggplot(compiled.PFHxS.SERUM, aes(x = log(Serum_PFHxS))) + geom_histogram(fill = "blue", binwidth = 0.2) + xlim(0,10) + ggtitle("PFHxS Serum Concentrations") 

ggplot(compiled.PFHxS.SERUM, aes(y = Serum_PFHxS)) +
  geom_boxplot(fill = "blue") +
  ggtitle("PFHxS Distribution") +
  geom_hline(yintercept = mean(compiled.PFHxS.SERUM$Serum_PFHxS), color = "red")


ggplot(compiled.PFHxS.SERUM, aes(y = Serum_PFHxS, x = GENDER, col = GENDER)) +
  geom_boxplot(fill = "blue") +
  ggtitle("PFHxS Distribution") +
  geom_hline(yintercept = mean(compiled.PFHxS.SERUM$Serum_PFHxS), color = "red") + facet_wrap(~GENDER)


ggplot(compiled.PFHxS.SERUM, aes(y = Serum_PFHxS)) + geom_boxplot(fill = "blue") + ggtitle("PFHxS Distribution")

ggplot(compiled.PFHxS.SERUM, aes(x = AGE, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs Age") + theme(plot.title = element_text(hjust = 0.5))
## Is this really a significant relationship? Why is it the most important?

ggplot(compiled.PFHxS.SERUM, aes(x = BMI, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs BMI") + theme(plot.title = element_text(hjust = 0.5))
## Outliers. Is the relationship legitimate or sampling bias? 

ggplot(compiled.PFHxS.SERUM, aes(x = URXUCR, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs URXUCR") + theme(plot.title = element_text(hjust = 0.5))
## Outliers. Is the relationship legitimate or sampling bias and outliers? 

ggplot(compiled.PFHxS.SERUM, aes(x = eGFR, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs eGFR") + theme(plot.title = element_text(hjust = 0.5))
## Outliers. Is the relationship legitimate or sampling bias and outliers? 

ggplot(compiled.PFHxS.SERUM, aes(x = LBXRDW, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs LBXRDW") + theme(plot.title = element_text(hjust = 0.5))
## Outliers. Is the relationship legitimate or sampling bias and outliers? 

ggplot(compiled.PFHxS.SERUM, aes(x = BMXHT, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs BMXHT") + theme(plot.title = element_text(hjust = 0.5))

ggplot(compiled.PFHxS.SERUM, aes(x = URXUMA, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs URXUMA") + theme(plot.title = element_text(hjust = 0.5))

ggplot(compiled.PFHxS.SERUM, aes(x = LBXLYPCT, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs LBXLYPCT") + theme(plot.title = element_text(hjust = 0.5))

ggplot(compiled.PFHxS.SERUM, aes(x = LBDRFO, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs LBDRFO") + theme(plot.title = element_text(hjust = 0.5))

ggplot(compiled.PFHxS.SERUM, aes(x = LBDFOT, y = Serum_PFHxS)) +
  geom_point(col = "blue") + stat_smooth(method = "lm", col = "red") + ggtitle("PFHxS Serum Concentration vs LBDFOT") + theme(plot.title = element_text(hjust = 0.5))
 ## See two significant outliers
```
:::

::: {style="margin-bottom:50px;"}
## Build a regression with the GBM variables

Variables identified as important:

AGE

BMI

URXUCR: Urinary creatinine (mg/dL)

eGFR

LBXRDW: Red cell distribution width (%)

BMXHT: Standing Height (cm)

URXUMA: Urinary albumin (ug/mL)

LBXLYPCT: Lymphocyte percent (Coulter) (%)

LBDRFO: RBC folate (ng/mL)

LBDFOT: Serum total folate (ng/mL)

```{r}

gbm.reg <- lm(Serum_PFHxS ~ AGE + BMI + URXUCR + eGFR + LBXRDW + BMXHT + URXUMA + LBXLYPCT + LBDRFO + LBDFOT , data = compiled.PFHxS.SERUM)

summary(gbm.reg)

qqnorm(gbm.reg$residuals)

qqline(gbm.reg$residuals)

```
:::

::: {style="margin-bottom:50px;"}
## Logarithmic Test Model

```{r}
gbm.reg <- lm(log(Serum_PFHxS) ~ AGE + BMI + URXUCR + eGFR + LBXRDW + BMXHT + URXUMA + LBXLYPCT + LBDRFO + LBDFOT , data = compiled.PFHxS.SERUM)

summary(gbm.reg)

qqnorm(gbm.reg$residuals)

qqline(gbm.reg$residuals)
```
:::
